Conditionally Autoregressive Models for House Price Data: Insights From a Comparative Simulation Study

### 2025-05-06 manuscript.qmd
update manuscript and response

# Study region
In the spatial modelling, the study region $\mathcal{S}$ is typically divided into $K$ distinct non-overlapping geographic units, denoted as $\mathcal{S}_k$ for $k = 1, 2, \dots, K$. Each geographic unit $\mathcal{S}_k$ is associated with a target variable $y_k$ and a set of explanatory factors represented as a vector $\mathbf{x}_k$.

@@

$$
\begin{array}{cccccc}
    & A & B & C & D & E \\
A & 0 & 1 & 1 & 1 & 1\\ 
B & 1 & 0 & 1 & 0 & 0\\ 
C & 1 & 1 & 0 & 1 & 0\\
D & 1 & 0 & 1 & 0 & 1\\
E & 1 & 0 & 0 & 1 & 0\\
\end{array}
$$

@@

For instance, in SAR and CAR models, the diagonal matrix $\boldsymbol{D}$ aids in incorporating spatial lag and error components by appropriately weighting the influence of neighbouring regions (@wall2004close, @ver2018spatial). In GWR, this matrix assists in locally calibrating the model by reflecting the density and connectivity of regions (@brunsdon1996geographically, @stewart2018localized). Here, $\boldsymbol{D}$ is represented as:

$$
\begin{array}{cccccc}
   & A & B & C & D & E \\
A & 4 & 0 & 0 & 0 & 0\\ 
B & 0 & 2 & 0 & 0 & 0\\ 
C & 0 & 0 & 3 & 0 & 0\\
D & 0 & 0 & 0 & 3 & 0\\
E & 0 & 0 & 0 & 0 & 2\\
\end{array}
$$

This matrix format indicates that region $A$ has 4 neighbour, regions $C$ and $D$ each have 3 neighbours, and region $B$ and $E$ has 2 neighbour.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#| label: fig-neighartfreg
#| fig-cap: "Neighbourhood structure of the artificial region. This figure illustrates the spatial connectivity between polygons in the artificial study region, where each line represents a defined neighbour relationship based on spatial adjacency"
#| out-width: "70%"
#| fig-align: "center"

@@

# Spatial regression model
When discussing spatial regression, it's crucial to comprehend the basic notion of linear regression. In classical linear regression, the relationship between the dependent variable $y$ and the $x_1,x_2, \dots, x_p$ independent variables is expressed as

$$
y = \beta_0 + \beta_1x_1 + \dots + \beta_p x_p + \varepsilon
$${#eq-basicreg}

@@

$$
y_i(s) = \beta_{0}(s) + \sum_{k=1}^p \beta_k(s)x_{ik}(s) + \varepsilon(s), \\ i = 1,\dots, n
$${#eq-gwr}

@@

$$
\beta(s) = (\boldsymbol{x}^T \boldsymbol{W}(s) \boldsymbol{x})^{-1} \boldsymbol{x}^T \boldsymbol{W}(s) \boldsymbol{y}
$${#eq-betagwr}

@@

## Simultaneously Autoregressive (SAR) Models
While GWR focuses on capturing localised spatial heterogeneity by allowing coefficients to vary across space, SAR take a different approach by explicitly modelling spatial dependencies through a spatial lag. The SAR model is a spatial econometric model used to analyse spatial dependencies and relationships among observations in a geographic space (@anselin1988spatial). It is widely employed in various fields, such as regional economics, environmental studies, and urban planning. SAR is a type of spatial autoregressive model involving a simultaneous equation framework to capture spatial interactions.

@@

$$
y(s) = \rho \sum_{s'} w(s,s') y(s') + \sum_{k=1}^p \beta_k x_k(s) + \varepsilon(s)
$${#eq-sar}

@@

$$
\mathbf{Y} \sim \mathcal{N}\left( \left( I - \rho W \right)^{-1} X\beta, \, \sigma^2 \left( I - \rho W \right)^{-1} \left( I - \rho W^T \right)^{-1} \right)
$${#eq-jointysar}

@@

$$
y(s) = \sum_{k=1}^p \beta_k x_k(s) + \varepsilon(s) + \phi(s) 
$${#eq-car}

Here, $\boldsymbol{\varepsilon} \sim N(0, \nu^2)$ while $\boldsymbol{\phi}$ is a specific component in CAR model that has a role as spatial effect. It is also common to mention $\boldsymbol{\phi}$ as a CAR priors. It is a type of Gaussian Markov random field (@rue2005gaussian), capture spatial autocorrelation by ensuring that values at nearby locations are more similar than those further apart (@lee2013carbayes). This can be expressed in a general term

$$
\boldsymbol{\phi}\sim N(0,\tau^2\boldsymbol{Q}^{-1}) 
$${#eq-phidist}

@@

$$
\phi_k|\boldsymbol\phi_{-K}, \boldsymbol{W},\rho, \tau^2  \sim  N\left(\frac{\rho\sum_{i=1}^Kw_{ki}\phi_i}{A},\frac{\tau^2}{A}\right)
$${#eq-phicar}

@@

$$
y_i(s) = \sum_{k=1}^p \beta_k x_{ki}(s) + \varepsilon_i(s) + \phi(s)
$${#eq-ycar}

The Bayesian approach to CAR models entails defining prior distributions for all model parameters, such as the regression coefficients $\beta$, variance parameters $\nu^2$ and $\tau^2$, and spatial random effects $\phi_k$. Subsequently, Bayesian inference techniques like Markov Chain Monte Carlo (MCMC) are employed to derive posterior distributions of these parameters. This method facilitates the integration of prior knowledge and offers a versatile framework for quantifying uncertainty.

@@

# add estimation method for each model
kbl(
  data.frame(
    Elements = c(
      "Number of Observations ($N$)", 
      "Covariate Matrix ($X$)", 
      "Spatial Weights Matrix ($W$)", 
      "Spatial Coordinates (coords)", 
      "Covariate Coefficients ($\\beta$)", 
      "Error Term variance ($\\sigma^2$)", 
      "Spatial Autocorrelation ($\\rho$)", 
      "Spatial Structure component ($\\phi$)", 
      "Spatial structure variance ($\\tau^2$)", 
      "Additional scale ($\\nu^2$)",
      "Estimation method"
    ),
    GWR = c("$\\checkmark$", "$\\checkmark$", "$\\times$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\times$", "$\\times$", "$\\times$", "$\\times$", "MLE"),
    SAR = c("$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\times$", "$\\times$", "$\\times$", "MLE/GMM"),
    CAR = c("$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "MCMC")
  ),
  col.names = c("Elements", "$GWR$", "$SAR$", "$CAR$"),
  escape = FALSE, # Needed to render LaTeX math symbols
  booktabs = TRUE,
  longtable = TRUE) %>% 
  kable_styling(font_size = 9)
```

@@

By employing artificially generated covariates and an artificially constructed study region, we developed three distinct property price datasets, each based on a different spatial model: CAR, a combination of CAR and SAR, and GWR. The true parameter values for this purpose are $\boldsymbol{\beta} = \begin{bmatrix} 1 & 0.9 & 0.7 & 0.5 & 0.3 \end{bmatrix}$, $\nu^2 = 0.2$, $\tau^2 = 0.6$ and $\rho = 0.6$. Whereas beta values for GWR dataset is generated using a beta function, which is a function of coordinate points. The matrix $\mathbf{W}$ was derived from the neighbourhood structure of an artificial study region using the functions `poly2nb()` and `nb2mat()`. Data generation for $SAR + CAR$ employs a slightly different strategy. In addition to being generated per area simultaneously using the joint distribution of $\mathbf{Y}$ as in @eq-jointysar, information about the spatial effect $\boldsymbol{\phi}$ is also added at the end of the process. Therefore, we refer to it as $SAR + CAR$ data.

@@

::: {#fig-biasplot layout-ncol=2}

![Bias values by method and dataset type for area-level data](figures/bias1_plot.png){#fig-biasplot1}

@@

Lombok includes five districts (kota/kabupaten), 53 sub-districts (kecamatan), and 608 villages (desa/kelurahan), which represent three successive levels of administrative division. In this study, the sub-district level is used as the spatial unit of analysis, as illustrated in @fig-lombokmap. The key variables of interest include property price, land size, built-up area, number of bedrooms, and number of bathrooms.

@@

Further, we conducted several preprocessing steps. First, we filtered the dataset by removing irrelevant variables, including floors, furnishing status, and category. It was then subsequently filtered to retain only properties with plausible characteristics: land area between 90 and 800 square meters, built-up size between 70 and 600 square meters, a maximum of 6 bedrooms, and a maximum 5 bathrooms. These thresholds were applied to ensure the data reflect realistic and context-appropriate housing characteristics in Lombok, based on common residential patterns and local housing norms. This step was essential to enhance the validity of the analysis by excluding outliers or potentially erroneous entries. Following this procedure, the dataset was refined to comprise a total of 576 observations. A summary of the key variables is presented in @tbl-datasumm.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#| label: tbl-datasumm
#| tbl-cap: "Summary statistics of variables in the property price dataset"
#| eval: true

# Create the data frame
data <- data.frame(
  Variable = c("Prices (million IDR)", "Land-size (sqm)", "Built-up area (sqm)", 
               "No. of bedroom", "No. of bathroom"),
  Mean   = round(c(1983, 299.6, 184.8, 3.24, 2.55), 2),
  Median = round(c(1400, 255.0, 150.0, 3.00, 2.00), 2),
  Q1     = round(c(800, 147.2, 100.0, 2.00, 2.00), 2),
  Q3     = round(c(2500, 407.2, 210.0, 4.00, 3.00), 2),
  Min    = round(c(170, 90.0, 70.0, 1.00, 1.00), 2),
  Max    = round(c(22500, 800.0, 600.0, 6.00, 5.00), 2),
  SD     = round(c(2056.81, 179.39, 109.20, 1.19, 1.05), 2),
  `NA`   = c(26, 24, 24, 24, 24)
)

# Display the table with knitr::kable
kbl(data, 
         col.names = c("Variable", "Mean", "Median", 
                       "$Q_1$", "$Q_3$", "Min", "Max", "SD", "NA"),
    escape = FALSE,
    booktabs = TRUE,
    longtable = TRUE) %>% 
  kable_styling(font_size = 9)
```

@@

To address missing values, we employed the `mice` package (@van2011mice) to perform multiple imputation with parameters set to $maxit = 10$, $m = 10$, and a fixed random seed to ensure reproducibility. The imputed values were then reintegrated into the main dataset, replacing the original missing entries. This process resulted in a final dataset that was complete and ready for subsequent analysis using the selected spatial models.

@@

@@

The cleaned and completed Lombok dataset was then model. The tables below present the results of the data fitting, offering a comparative view of each model’s effectiveness and accuracy in capturing the dataset's characteristics.

$$
prices \sim land_{std} + building_{std} + beds_{std} + baths_{std}
$${#eq-lopreg}

Standardization of the predictors ensures that the resulting coefficients are on a comparable scale, facilitating meaningful interpretation and comparison across different modeling approaches. The results presented in @tbl-modelcomp provide a comprehensive evaluation of each model’s performance in capturing both the structural and spatial characteristics of the property market in Lombok.

We conducted model fitting on the area-level data, with detailed results provided in the Appendix (@tbl-modelcomp1). However, this method does not provide optimal parameter estimates because data aggregation can reduce variability and obscure spatial details essential for capturing local effects. To address this, we used point-level data to apply three different models: GLMM, GWR, and multilevel CAR.

@@

# Define the dataset using tribble
table_data <- tribble(
  ~"",                             ~GLMM,                    ~GWR,                     ~mlvCAR,
  "Intercept",                   "7.09 [6.93, 7.24]*",     "7.25 [7.25, 7.26]*",     "7.08 [6.99, 7.17]*",
  "Land-size",                   "0.26 [0.21, 0.32]*",     "0.31 [0.30, 0.32]*",     "0.26 [0.21, 0.32]*",
  "Built-up area",               "0.16 [0.10, 0.22]*",     "0.19 [0.19, 0.20]*",     "0.16 [0.10, 0.21]*",
  "No. of bedroom",              "-0.03 [-0.09, 0.01]",    "-0.15 [-0.17, -0.15]*",  "-0.03 [-0.09, 0.02]",
  "No. of bathroom",             "0.19 [0.13, 0.25]*",     "0.24 [0.23, 0.26]*",     "0.19 [0.13, 0.25]*",
  "$\\rho$",                     "-",                      "-",                      "0.48 [0.07, 0.90]*",
  "$\\nu^2$",                    "0.22 [0.19, 0.25]*",     "0.33 [0.29, 0.37]*",     "0.22 [0.20, 0.25]*",
  "$\\tau^2$",                   "0.20 [0.11, 0.34]*",     "-",                      "0.31 [0.14, 0.58]*",
  "AIC",                         "803.38",                 "-645.104",               "786.95",
  "DIC",                         "-",                      "-",                      "850.39",
  "WAIC",                        "838.3",                  "-",                      "851.14",
  "LMPL",                        "-421.31",               "-",                      "-427.21",
  "Log-likelihood",              "-400.69",                "-520.97",                "-385.47"
)

@@

Here, @tbl-modelcomp compares parameter estimates and fit criteria for three models: GLMM, GWR, and multilevel CAR. Significant coefficients are marked with an asterisk (*). Key determinants like land-size and built-up area have positive, significant effects across models, while the number of bedrooms shows only minor, varying effects, indicating model-specific differences in capturing this variable's influence. Only the multilevel CAR model includes the spatial parameter $\rho$, with a significant estimate of 0.48, showing the model's ability to account for spatial dependence. Variance components $\nu^2$ (residual) and $\tau^2$ (spatial) are also detailed, with both variance terms significant. Fit criteria (AIC, WAIC, and Log-likelihood) suggest mlvCAR and GWR may better capture local spatial variation, with mlvCAR balancing spatial structure and precision.

::: {#fig-pricenphi layout-ncol=2}

![phi values](figures/phi_map.png){#fig-pricenphi1}

@@

@fig-pricenphi1 and @fig-pricenphi2 respectively present spatial heatmaps illustrating the distribution of estimated spatial random effects ($\phi$) from the CAR model and the average log-transformed house prices per sub-district accross Lombok island.

Here, the $\phi$ values represent the spatially structured residuals not explained by the observed covariates in the CAR model. Specifically, it captures latent spatial influences such as neighbourhood desirability, access to infrastructure, or even local policy, that affect house prices beyond what structural features like land and building size can account for.

Because the model uses log-transformed prices, $\phi$ has multiplicative interpretation on the original price scale. For instance, if $\phi = 0.57$ for a given sub-district, then after accounting for covariates, properties in that area are estimated to be approximately $exp(0.57) \approx 1.76$ or 76% more expensive than average. Similarly, a sub-district with $\phi = -0.2$, this suggests that properties in that area are approximately 18% less expensive than average, since $exp(-0.2) \approx 0.82$.

@@

The results underscore the importance of using models that explicitly account for spatial dependencies in fields like property valuation, where spatial heterogeneity plays a significant role. The superior performance of the CAR model, particularly in datasets with strong spatial structure, suggests that it may be a more reliable choice for applications where local spatial variations are critical. The findings align with previous studies indicating that CAR models are better suited for managing spatial autocorrelation and heterogeneity, especially when spatial dependencies are strong.

@@

These findings echo the limitations of GWR when precise coordinates are lacking, affirming the relevance of area-level approaches. Moreover, while we use a binary adjacency matrix for consistency, we recognise that the choice of spatial weight $\mathbf{W}$ may influences model performance. Future studies should explore alternative specifications such as row-standardized or distance-decay matrices to assess their impact on robustness and interpretability.

@@

* Funding
* Competing interests

@@

* Ethics approval
* Consent to participate
* Consent for publication
* Availability of data and materials
* Code availability
* Authors' contributions

# Appendix {.unnumbered}

@@

## Model Comparison for area-level data{.unnumbered}

@@

## Example of beta-values in GWR model{.unnumbered}

>>>>

# Study region

In the spatial modelling, the study region $\mathcal{S}$ is typically divided into $K$ distinct non-overlapping geographic units, denoted as $\mathcal{S}_k$ for $k = 1, 2, \dots, K$. Each geographic unit $\mathcal{S}_k$ is associated with a target variable $y_k$ and a set of explanatory factors represented as a vector $\mathbf{x}_k$.

@@

$$
\mathbf{W} =
\begin{array}{cccccc}
   & A & B & C & D & E \\
A & 0 & 1 & 1 & 1 & 1\\ 
B & 1 & 0 & 1 & 0 & 0\\ 
C & 1 & 1 & 0 & 1 & 0\\
D & 1 & 0 & 1 & 0 & 1\\
E & 1 & 0 & 0 & 1 & 0\\
\end{array}
\hspace{2.5cm}
\mathbf{D} =
\begin{array}{cccccc}
   & A & B & C & D & E \\
A & 4 & 0 & 0 & 0 & 0\\ 
B & 0 & 2 & 0 & 0 & 0\\ 
C & 0 & 0 & 3 & 0 & 0\\
D & 0 & 0 & 0 & 3 & 0\\
E & 0 & 0 & 0 & 0 & 2\\
\end{array}
$$

@@

For instance, in SAR and CAR models, the diagonal matrix $\boldsymbol{D}$ aids in incorporating spatial lag and error components by appropriately weighting the influence of neighbouring regions (@wall2004close, @ver2018spatial). In GWR, this matrix assists in locally calibrating the model by reflecting the density and connectivity of regions (@brunsdon1996geographically, @stewart2018localized). In this case, $\mathbf{D}$ indicates that region $A$ has 4 neighbours, regions $C$ and $D$ each have 3 neighbours, and region $B$ and $E$ has 2 neighbours.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#| label: fig-neighartfreg
#| fig-cap: "Neighbourhood structure of the artificial region. This figure illustrates the spatial connectivity between polygons in the artificial study region, where each line represents a defined neighbour relationship based on spatial adjacency"
#| out-width: "60%"
#| fig-align: "center"

@@

# Spatial regression model

When discussing spatial regression, it's crucial to comprehend the basic notion of linear regression. In classical linear regression, the relationship between the dependent variable $y$ and the $x_1,x_2, \dots, x_p$ independent variables is expressed as

$$
y = \beta_0 + \beta_1x_1 + \dots + \beta_p x_p + \varepsilon
$$ {#eq-basicreg}

@@

$$
y_i(s) = \beta_{0}(s) + \sum_{k=1}^p \beta_k(s)x_{ik}(s) + \varepsilon(s), \\ i = 1,\dots, n
$$ {#eq-gwr}

@@

$$
\beta(s) = (\boldsymbol{x}^T \boldsymbol{W}(s) \boldsymbol{x})^{-1} \boldsymbol{x}^T \boldsymbol{W}(s) \boldsymbol{y}
$$ {#eq-betagwr}

@@

## Simultaneously Autoregressive (SAR) Models

While GWR focuses on capturing localised spatial heterogeneity by allowing coefficients to vary across space, SAR take a different approach by explicitly modelling spatial dependencies through a spatial lag. The SAR model is a spatial econometric model used to analyse spatial dependencies and relationships among observations in a geographic space (@anselin1988spatial). It is widely employed in various fields, such as regional economics, environmental studies, and urban planning. SAR is a type of spatial autoregressive model involving a simultaneous equation framework to capture spatial interactions.

@@

$$
y(s) = \rho \sum_{s'} w(s,s') y(s') + \sum_{k=1}^p \beta_k x_k(s) + \varepsilon(s)
$$ {#eq-sar}

@@

$$
\mathbf{Y} \sim \mathcal{N}\left( \left( I - \rho W \right)^{-1} X\beta, \, \sigma^2 \left( I - \rho W \right)^{-1} \left( I - \rho W^T \right)^{-1} \right)
$$ {#eq-jointysar}

@@

$$
y(s) = \sum_{k=1}^p \beta_k x_k(s) + \varepsilon(s) + \phi(s) 
$$ {#eq-car}

Here, $\boldsymbol{\varepsilon} \sim N(0, \sigma_{\varepsilon}^2)$ while $\boldsymbol{\phi}$ is a specific component in CAR model that has a role as spatial effect. It is also common to mention $\boldsymbol{\phi}$ as a CAR priors. It is a type of Gaussian Markov random field (@rue2005gaussian), capture spatial autocorrelation by ensuring that values at nearby locations are more similar than those further apart (@lee2013carbayes). This can be expressed in a general term

$$
\boldsymbol{\phi}\sim N(0,\sigma_{\phi}^2\boldsymbol{Q}^{-1}) 
$$ {#eq-phidist}

@@

$$
\phi_k|\boldsymbol{\phi}_{-k}, \mathbf{W},\rho, \sigma_{\phi}^2  \sim  N\left(\frac{\rho\sum_{i=1}^Kw_{ki}\phi_i}{A},\frac{\sigma_{\phi}^2}{A}\right)
$$ {#eq-phicar}

@@

$$
y_i(s) = \sum_{k=1}^p \beta_k x_{ki}(s) + \varepsilon_i(s) + \phi(s)
$$ {#eq-ycar}

The Bayesian approach to CAR models entails defining prior distributions for all model parameters, such as the regression coefficients $\beta$, variance parameters $\sigma_{\varepsilon}^2$ and $\sigma_{\phi}^2$, and spatial random effects $\phi_k$. Subsequently, Bayesian inference techniques like Markov Chain Monte Carlo (MCMC) are employed to derive posterior distributions of these parameters. This method facilitates the integration of prior knowledge and offers a versatile framework for quantifying uncertainty.

@@

# add estimation method for each model
kbl(
  data.frame(
    Elements = c(
      "Number of Observations ($N$)", 
      "Covariate Matrix ($X$)", 
      "Spatial Weights Matrix ($W$)", 
      "Spatial Coordinates (coords)", 
      "Covariate Coefficients ($\\beta$)", 
      "Spatial Autocorrelation ($\\rho$)", 
      "Spatial Structure component ($\\phi$)", 
      "Spatial structure variance ($\\sigma_{\\phi}^2$)", 
      "Error Term variance ($\\sigma_{\\varepsilon}^2$)",
      "Estimation method"
    ),
    GWR = c("$\\checkmark$", "$\\checkmark$", "$\\times$", "$\\checkmark$", "$\\checkmark$", "$\\times$", "$\\times$", "$\\times$", "$\\checkmark$", "MLE"),
    SAR = c("$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\times$", "$\\times$", "$\\checkmark$", "MLE/GMM"),
    CAR = c("$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$",  "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "MCMC")
  ),
  col.names = c("Elements", "$GWR$", "$SAR$", "$CAR$"),
  escape = FALSE, # Needed to render LaTeX math symbols
  booktabs = TRUE,
  longtable = TRUE) %>% 
  kable_styling(font_size = 9)
```

@@

By employing artificially generated covariates and an artificially constructed study region, we developed three distinct property price datasets, each based on a different spatial model: CAR, a combination of CAR and SAR, and GWR. The true parameter values for this purpose are $\boldsymbol{\beta} = \begin{bmatrix} 1 & 0.9 & 0.7 & 0.5 & 0.3 \end{bmatrix}$, $\sigma_{\phi}^2 = 0.2$, $\sigma_{\phi}^2 = 0.6$ and $\rho = 0.6$. Whereas beta values for GWR dataset is generated using a beta function, which is a function of coordinate points. The matrix $\mathbf{W}$ was derived from the neighbourhood structure of an artificial study region using the functions `poly2nb()` and `nb2mat()`. Data generation for $SAR + CAR$ employs a slightly different strategy. In addition to being generated per area simultaneously using the joint distribution of $\mathbf{Y}$ as in @eq-jointysar, information about the spatial effect $\boldsymbol{\phi}$ is also added at the end of the process. Therefore, we refer to it as $SAR + CAR$ data.

@@

::: {#fig-biasplot layout-ncol="2"}
![Bias values by method and dataset type for area-level data](figures/bias1_plot.png){#fig-biasplot1}

@@

Lombok comprises five districts (kota/kabupaten), 53 sub-districts (kecamatan), and 608 villages (desa/kelurahan), representing three successive levels of administrative division. Mataram District, despite its small area compared to other districts in Lombok, is the most urbanized and economically active area on the island, characterized by dense residential clusters with relatively homogeneous housing specifications. In adjacent districts such as West Lombok and Central Lombok, housing developments have expanded, particularly to accommodate commuters working in Mataram. The coastal areas of West Lombok, including Batu Layar and Sekotong, exhibit villa-style housing commonly associated with tourism, resulting in generally higher property prices. A similar pattern is found in North Lombok, where mountainous terrain limits development, but tourism-driven demand has led to villa-style housing near popular destinations. East Lombok, by contrast, remains more isolated due to its distance from Mataram. Housing in this district tends to serve primarily the local population, with limited signs of external investment or tourism-driven development. Additionally, the southern part of Central Lombok (Pujut) has seen emerging residential and real estate activity following the construction of the international MotoGP circuit.

In this study, the sub-district level is used as the spatial unit of analysis, as illustrated in @fig-lombokmap. The key variables of interest include property price, land size, built-up area, number of bedrooms, and number of bathrooms.

@@

Further, we conducted several preprocessing steps. First, we filtered the dataset by removing irrelevant variables, including floors, furnishing status, and category. It was then subsequently filtered to retain only properties with plausible characteristics: land area between 90 and 800 square meters, built-up size between 70 and 600 square meters, a maximum of 6 bedrooms, and a maximum 5 bathrooms. These thresholds were applied to ensure the data reflect realistic and context-appropriate housing characteristics in Lombok, based on common residential patterns and local housing norms. This step was essential to enhance the validity of the analysis by excluding outliers or potentially erroneous entries. Following this procedure, the dataset was refined to comprise a total of 576 observations. A summary of the key variables is presented in @tbl-datasumm. Notably, the distribution of property prices is skewed and does not exhibit a bell-shaped curve, which motivates the inclusion of a log-transformed price variable in the analysis.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#| label: tbl-datasumm
#| tbl-cap: "Summary statistics of house prices dataset ($N = 576$, `NA` = 4%), including variables such as prices and its logscale value, land area(sqm), building area(sqm), number of beds, and number of baths"
#| eval: true

load("data/propdata.RData")
load("data/hp_dat_v1.RData")

propdata <- propdata %>%
  select(-village, -floors, -furnished, -category) %>%
  rbind(hp_dat_v1) %>%
  filter(land >= 90 & land <= 800) %>%
  filter(building >= 70 & building <= 600) %>%
  filter(beds <= 6) %>%
  filter(baths > 0 & baths <= 5) %>% 
  mutate(
    prices = prices * 1000,
    logprices = log(prices)
    )

dat <- propdata %>% 
  select(prices, logprices, land, building, beds, baths)

# Create the data frame
data <- data.frame(
  Variable = c("Prices (million IDR)", "logPrices", "Land-size (sqm)", "Built-up area (sqm)",
               "No. of bedroom", "No. of bathroom"),
  Mean   = round(colMeans(dat, na.rm = TRUE), 2),
  SD     = round(apply(dat, 2, sd, na.rm = TRUE), 2),
  Median = round(apply(dat, 2, median, na.rm = TRUE), 2),
  Q1     = round(apply(dat, 2, quantile, probs = 0.25, na.rm = TRUE), 2),
  Q3     = round(apply(dat, 2, quantile, probs = 0.75, na.rm = TRUE), 2),
  Min    = round(apply(dat, 2, min, na.rm = TRUE), 2),
  Max    = round(apply(dat, 2, max, na.rm = TRUE), 2)
)

# Display the table with knitr::kable
kbl(data,
    col.names = c("Variable", "Mean", "SD", "Median", "$Q_1$", "$Q_3$", "Min", "Max"),
    row.names = FALSE,
    escape = FALSE,
    booktabs = TRUE,
    longtable = TRUE) %>%
  kable_styling(font_size = 9)
```

@@

To address missing values, we used the `mice` package (@van2011mice), which performs multiple imputation by chained equations (MICE). This approach iteratively fills in missing data by modeling each variable with missing values as a function of other variables in the dataset. By generating multiple imputed datasets, it accounts for the uncertainty inherent in missing data. In our case, we specified 10 iterations and generated 10 imputed datasets. A fixed random seed was used to ensure the results were reproducible. The imputed datasets were then pooled to obtain final estimates for subsequent analyses. The imputed values were then reintegrated into the main dataset, replacing the original missing entries. This process resulted in a final dataset that was complete and ready for subsequent analysis using the selected spatial models.

@@

Moreover, as GWR requires spatial coordinates for each observation, we addressed this by generating random points around the centroid of each sub-district using the `st_sample()` function. The number of points generated matched the number of observations in each area, and each was positioned within a 0.5 km radius of the centroid to maintain a realistic spatial distribution while preserving the local context. This strategy allowed us to simulate plausible spatial locations for property transactions in the absence of precise geolocation data. Importantly, we applied the same bandwidth in the GWR model fitting to ensure consistency. By preserving the area-level grouping structure, this approach supports a fair comparison with the previous method.

@@

The cleaned and completed Lombok dataset was then fitted to the model. Equation @eq-lopreg presents the proposed model specification, which serves as the fixed formula applied consistently across all analyses

$$
\text{logprices} \sim \text{land}_{\text{std}} + \text{building}_{\text{std}} + \text{beds}_{\text{std}} + \text{baths}_{\text{std}}
$$ {#eq-lopreg}

Standardization of the predictors ensures that the resulting coefficients are on a comparable scale, facilitating meaningful interpretation and comparison across different modeling approaches. We conducted model fitting on the area-level data, with detailed results provided in the Appendix (@tbl-modelcomp1). However, this method does not provide optimal parameter estimates because data aggregation can reduce variability and obscure spatial details essential for capturing local effects. To address this, we used point-level data to apply three different models: GLMM, GWR, and multilevel CAR.

The results presented in @tbl-modelcomp provide a comprehensive evaluation of each model’s performance in capturing both the structural and spatial characteristics of the property market in Lombok.

@@

# Define the dataset using tribble
table_data <- tribble(
  ~"",                             ~GLMM,                    ~GWR,                     ~mlvCAR,
  "Intercept",                   "7.09 [6.93, 7.24]*",     "7.25 [7.25, 7.26]*",     "7.08 [6.99, 7.17]*",
  "Land-size",                   "0.26 [0.21, 0.32]*",     "0.31 [0.30, 0.32]*",     "0.26 [0.21, 0.32]*",
  "Built-up area",               "0.16 [0.10, 0.22]*",     "0.19 [0.19, 0.20]*",     "0.16 [0.10, 0.21]*",
  "No. of bedroom",              "-0.03 [-0.09, 0.01]",    "-0.15 [-0.17, -0.15]*",  "-0.03 [-0.09, 0.02]",
  "No. of bathroom",             "0.19 [0.13, 0.25]*",     "0.24 [0.23, 0.26]*",     "0.19 [0.13, 0.25]*",
  "$\\rho$",                     "-",                      "-",                      "0.48 [0.07, 0.90]*",
  "$\\sigma_{\\varepsilon}^2$",  "0.22 [0.19, 0.25]*",     "0.33 [0.29, 0.37]*",     "0.22 [0.20, 0.25]*",
  "$\\sigma_{\\phi}^2$",          "0.20 [0.11, 0.34]*",     "-",                      "0.31 [0.14, 0.58]*",
  "AIC",                         "803.38",                 "-645.104",               "786.95",
  "DIC",                         "-",                      "-",                      "850.39",
  "WAIC",                        "838.3",                  "-",                      "851.14",
  "LMPL",                        "-421.31",               "-",                      "-427.21",
  "Log-likelihood",              "-400.69",                "-520.97",                "-385.47"
)

@@

Here, @tbl-modelcomp compares parameter estimates and fit criteria for three models: GLMM, GWR, and multilevel CAR. Significant coefficients are marked with an asterisk (\*). Key determinants like land-size and built-up area have positive, significant effects across models, while the number of bedrooms shows only minor, varying effects, indicating model-specific differences in capturing this variable's influence. Only the multilevel CAR model includes the spatial parameter $\rho$, with a significant estimate of 0.48, showing the model's ability to account for spatial dependence. Variance components $\sigma_{\varepsilon}^2$ (residual) and $\sigma_{\phi}^2$ are also detailed, with both terms are significant. Fit criteria (AIC, WAIC, and Log-likelihood) suggest mlvCAR and GWR may better capture local spatial variation, with mlvCAR balancing spatial structure and precision.

::: {#fig-pricenphi layout-ncol="2"}
![$\phi$ values](figures/phi_map.png){#fig-pricenphi1}

@@

@fig-pricenphi1 and @fig-pricenphi2 respectively present spatial heatmaps illustrating the distribution of estimated spatial random effects ($\phi$) from the CAR model and the average log-transformed house prices per sub-district accross Lombok island. It can be seen that both plots in @fig-pricenphi exhibit similar spatial patterns, with high and low values clustering in comparable regions. Note that they do not have to be identical, as $\phi$ represents the spatially structured residual component after accounting for covariates, whereas the log-transformed prices reflect the observed aggregated market values. Nonetheless, their resemblance suggests that some of the spatial variability in house prices is being effectively captured by the spatial random effects.

Because the model uses log-transformed prices, $\phi$ has multiplicative interpretation on the original price scale. For instance, if $\phi = 0.57$ for a given sub-district, then after accounting for covariates, properties in that area are estimated to be approximately $\exp(0.57) \approx 1.76$ or 76% more expensive than average. Similarly, a sub-district with $\phi = -0.2$, this suggests that properties in that area are approximately 18% less expensive than average, since $\exp(-0.2) \approx 0.82$. Note that the 95% credible interval of $\phi$ can be interpreted similarly by exponentiating the lower and upper bounds, yielding a multiplicative uncertainty range for the price deviation.

@@

The results underscore the importance of using models that explicitly account for spatial dependencies in fields like property valuation, where spatial heterogeneity plays a significant role. The superior performance of the CAR model, particularly in datasets with strong spatial structure, suggests that it may be a more reliable choice for applications where local spatial variations are critical. The findings align with previous studies (@Wang2013A, @Soroori2019Application, @Zeng2014Bayesian, @Guelat2018Effects) indicating that CAR models are better suited for managing spatial autocorrelation and heterogeneity, especially when spatial dependencies are strong.

@@

These findings also echo the limitations of GWR when precise coordinates are lacking, affirming the relevance of area-level approaches. Moreover, while we use a binary adjacency matrix for consistency, we recognise that the choice of spatial weight $\mathbf{W}$ may influences model performance. Future studies should explore alternative specifications such as row-standardized or distance-decay matrices to assess their impact on robustness and interpretability.

@@

-   Funding
-   Competing interests

@@

-   Ethics approval
-   Consent to participate
-   Consent for publication
-   Availability of data and materials
-   Code availability
-   Authors' contributions

\newpage
# Appendix {.unnumbered}

@@

## Model Comparison for area-level data {.unnumbered}

@@

## Example of beta-values in GWR model {.unnumbered}
